Trends from 30-year observations of downward solar irradiance in Thessaloniki, Greece

### 2023-12-20 MDPI_submission.tex
Commit 2023-12-20 11:14

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{The shortwave downward solar irradiance (SDR) is an important
factor that drives climate processes, production and can affect all
living organisms. Observations of SDR at different locations around the
world with different environmental characteristics have been used to
investigate its long-term variability and trends at different time
scales. Periods of positive trends are referred as brightening periods
and of negative trends as dimming periods. Here we studied 30 years of
pyranometer data in Thessaloniki, Greece, under three types of sky
conditions (clear sky, cloudy sky and all sky). The clear-sky data were
identified by applying a cloud screening algorithm. We found a positive
trend of \(0.38\,\%/\text{year}\) for all-sky, \(0.35\,\%/\text{year}\)
for clear-sky conditions, and \(-0.28\,\%/\text{year}\) for cloudy
conditions. We have also investigated the consistency of these trends,
their seasonal variability, and the effect of the solar zenith angle. We
have found that for all-sky and clear-sky conditions the SDR trend is
positive in winter (\(0.7\) and \(0.8\,\%/\text{year}\), respectively)
and autumn (\(~0.4\,\%/\text{year}\)), while under cloudy skies the
trend is negative (\(-0.9\,\%/\text{year}\) in winter and
\(-0.4\,\%/\text{year}\) in autumn). In spring and summer the trend is
very close to zero, irrespective of sky conditions. The SDR trend is
increasing with increasing solar zenith angle, except under cloudy skies
where the trend is highly variable and close to zero. Finally, we
identified some anomalies in the long term SDR trends for all sky
conditions by examining the cumulative sums of monthly anomalies from
the climatological mean.}

@@

The shortwave downward solar irradiance (SDR) at Earth's surface plays a
significant role, on its climate. Changes of the SDR can be related to
changes on Earth's energy budget, the mechanisms of climate change, and
water and carbon cycles \citep{Wild2009}. It can also affect solar and
agricultural production, and all living organisms. Studies of SDR
variability, have identified some distinct SDR trends on different
regions of the world on different time periods. The term `brightening'
is generally used to describe periods of positive SDR trend, and
`dimming' for negative trend \citep{Wild2009}. There are many cases in
the long term records of irradiance, showing a systematic change in the
magnitude of the trend, occurring roughly in the last decades of the
20th century. On multiple stations in China, a dimming period was
reported until about 2000, followed by a brightening period
\citep{Yang2021}. A similar pattern was identified, with the breaking
point around 1980, for stations in Central Europe \citep{Wild2021} and
Brazil \citep{Yamasoe2021}. On global scale, an artificial Intelligence
aided spatial analysis on continental level with data from multiple
stations reach similar conclusions for these regions and for the global
trend \citep{Yuan2021}.

@@

Due to the significant spatial and temporal variability of the trends,
and the contributing factors, there is a constant need to monitor and
investigate SDR in different sites in order to estimate the degree of
variability, and its relation to the local conditions. In this study, we
examine the trends of SDR, using ground-based measurements at
Thessaloniki, Greece, for the period 1993 to 2023, as derived from a
CM-21 pyranometer. We reevaluated and extended the dataset used by
\citet{Bais2013}, we applied a different algorithm for the
identification of clear-/cloud-sky instances
\citep{Reno2016, Reno2012a}, and we derived the SDR trends for the
period of study, under different sky conditions (all-sky, clear-sky and
cloud-sky). Finally, we investigated the dependence of the trends on
solar zenith angle and season.

@@

The SDR data were measured with a Kipp \& Zonen CM-21 pyranometer
operating continuously at the Laboratory of Atmospheric Physics of the
Aristotle University of Thessaloniki (\(40^\circ\,38'\,\)N,
\(22^\circ\,57'\,\)E, \(80\,\)m~a.s.l.) since 1993. Here we use data for
the period from 1993-04-13 to 2023-04-13. The monitoring site is located
near the city centre thus we expect that measurements are affected by
the urban environment. During the study period, the pyranometer has been
independently calibrated three times at the Meteorologisches
Observatorium Lindenberg, DWD, verifying that stability of the
instrument's sensitivity is better than \(0.7\%\) relative to the
initial calibration by the manufacturer. Along with SDR, the direct beam
radiation (DNI) is also measured with a collocated Kipp \& Zonen CHP-1
pyrheliometer since 2016-04-01. The DNI data were used as auxiliary data
to support the selection of appropriate thresholds in the clear sky
identification algorithm (CSid), which is discussed later. It is noted
that the limited dataset of DNI was not used for the identification of
clear sky cases in the entire SDR series to avoid any selection bias due
to unequal length of the two datasets. There are four distinct steps in
the creation of the dataset analyzed here: a)~the acquisition of
radiation measurements from the sensors, b)~the data quality check,
c)~the identification of ``clear sky'' conditions from the radiometric
data, and d)~the aggregation of data and trend analysis.

@@

In order to estimate the effect of the sky conditions on the long term
variability of SDR, we created three datasets by characterizing each
one-minute measurement with a corresponding sky-condition flag (i.e.,
all-sky, clear-sky and cloudy-sky). To identify the clear-cases we used
the method proposed by \citet{Reno2016}, which requires the definition
of some site specific parameters. These parameters, were produced by an
iterative process, as the original authors proposed and it are discussed
in the next section.

We note that all methods have some subjectivity in the definition of
clear or cloudy sky cases. As a result, the details of the definition
are site specific and they rely on a combination of thresholds and
comparisons with ideal radiation models and statistical analysis of
different signal metrics. The CSid algorithm was calibrated with the
main focus, to identify the presence of clouds. Despite the fine-tuning
of the procedure, in a few marginal cases false positive or false
negative results were identified by manual inspection. However, due to
their small number, they cannot affect the final results of the study.
For completeness, we provide below a brief overview of the clear-sky
identification algorithm (CSid), along with the site specific
thresholds.

@@

The linear trends were calculated using an first order autoregressive
model with lag of 1 day with fitting method of the ``maximum
likelihood'' \citep{Gardner1980, Jones1980}, by implementing the
function ``arima'' from the library ``stats'' of the R programming
language \citep{RCT2023}. The trends were reported together with the
\(2\sigma\) errors (Table~\ref{tab:trendtable}).

@@

We calculated the linear trends of SDR, from the departures of the mean
daily values from the daily climatology and for the three sky conditions
(Table~\ref{tab:trendtable}). In Figure~\ref{fig:trendALL} we present
only the time series under all-sky conditions; the plots for clear-sky
and cloud-sky conditions, are shown in the Appendix
(Figures~\ref{fig:trendCLEAR} and~ \ref{fig:trendCLOUD}). In the studied
period, there is no significant break or change in the variability
pattern of the time series. The linear trends in all three datasets are
positive and around \(0.4\,\%/y\) for all-sky and cloudy-sky conditions,
while for clear-skies the trend is much smaller
(\textasciitilde{}\(0.1\,\%/y\)). The linear trends were calculated
taking into account the autocorrelation of the time series and all three
are statistically significant at least at the \(95\,\%\) confidence
level, as they are larger than the corresponding \(2\sigma\) errors. The
clear-sky trend is very small suggesting a small effect from aerosols
and water vapor which are the dominant factors of the SDR variability.
In contrast the large positive trend of SDR under cloudy skies can be
attributed to reduction in cloud cover and/or cloud optical depth. Lack
of continuous cloud observations that could support these findings does
not allow drawing firm conclusions. However, there are indications that
the total cloud-cover as inferred from the ERA5 analysis for the grid
point of Thessaloniki is decreasing over the period of study.

@@

Although the year-to-year variability of the anomalies (Figure
\ref{fig:trendALL} and Figures~\ref{fig:trendCLEAR},
\ref{fig:trendCLOUD} in Appendix), shows a rather homogeneous behaviour,
plots of the cumulative sums (CUSUM) \citep{Regier2019} of the anomalies
can reveal different structures in the records of all three sky
conditions. For time series with a uniform trend, we would expect the
CUSUMs of the anomalies to have a symmetric `V' shape centered around
the middle of the series. This would indicate that the anomalies are
evenly distributed around the climatological mean, and for a positive
uniform trend, the first half is below and the second half above the
climatological mean. In our case, there is a more complex evolution of
the anomalies. For all-skies (Figure~\ref{fig:cusummonth-1}), we observe
three rather distinct periods: a) a downward part between the start of
the datasets and about 2000, denoting that all anomalies are negative
thus below the climatology; b) a relatively steady part lasting for
almost 20 years suggesting little variability in SDR anomalies; and c) a
steep upward part to present indicating anomalies above the climatology.
The CUSUMs for cloudy-skies (Figure~\ref{fig:cusummonth-3}), show a
similar behavior with some short-term differences that do not change the
overall pattern. For clear skies (Figure~\ref{fig:cusummonth-2}), a
monotonic downward tendency is evident until 2004, suggesting that the
anomalies are all negative. After 2004 the anomalies turn to positive at
a fast rate for about five years and at a slower rate thereafter.

@@

The trends under clear- and cloudy-sky conditions are of the same
direction and it would be expected that their sum is similar to the
all-sky trend. This does not happen, especially for winter, likely due
to the way the monthly means for clear and cloudy skies were calculated.
Daily means were calculated only when at least \(60\,\%\) of the clear-
or cloudy-sky data were available (see sect.
\ref{aggregationstatistical}).

@@

We have analyzed a 30 year dataset of SRD measurements in Thessaloniki
Greece (1993 -- 2023) aiming to identify the long term variability of
solar irradiance under different sky conditions. Under all-sky
conditions there is a positive trend in SDR of \(0.38\,\%/y\)
(brightening). A previous study \citep{Bais2013} for the period 1993 --
2011 reported also a positive trend, of \(0.33\,\%/y\). The slight
increase of this trend indicates that the brightening of SDR continues
and is probably caused by continuing decreases in aerosol optical depth
and the optical thickness of clouds over the area. A smaller trend has
been found under clear-sky conditions (\(0.097\,\%/y\)) which supports
the notion that part of the brightening is caused by decreasing
aerosols. \citet{Siomos2020} have shown that aerosol optical depth over
Thessaloniki is decreasing constantly at least up to 2018.
Unfortunately, for this study aerosol data for the entire period were
not available in order to quantify their effect on SDR. The attenuation
of SDR by aerosols over Europe has been proposed as major factor by
\citet{Wild2021}. The brightening effect on SDR under cloudy-sky
conditions (\(0.41\,\%/y\)), suggests that cloud optical thickness is
also decreasing during this period. As long term data of cloud optical
thickness are also not available for the region, we cannot draw
quantitative conclusions. \citet{Lozano2023} reported a Cloud Radiative
Forcing (CRF) trend of \(1.22\,W/m^2/y\) for the SDR for the city of
Granada Spain (\(37.16^\circ\,\)N, \(3.61^\circ\,\)E,
\(680\,\)m~a.s.l.). We found a comparable result, for the same
parameter, using the long term trend of all- and clear-skies, as follows
\(CRF = 1.46 - 0.501 = 0.959\,W/m^2/y\).

>>>>

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{The shortwave downward solar irradiance (SDR) is an important
factor that drives climate processes, energy production and can affect
all living organisms. Observations of SDR at different locations around
the world with different environmental characteristics have been used to
investigate its long-term variability and trends at different time
scales. Periods of positive trends are referred as brightening periods
and of negative trends as dimming periods. In this study we have used 30
years of pyranometer data in Thessaloniki, Greece, to investigate the
variability of SDR under three types of sky conditions (clear-, cloudy-
and all-sky). The clear-sky data were identified by applying a cloud
screening algorithm. We have found a positive trend of
\(0.38\,\%/\text{year}\) for all-sky, \(\sim 0.1\,\%/\text{year}\) for
clear-sky, and \(-0.41\,\%/\text{year}\) for cloudy conditions. The
consistency of these trends, their seasonal variability, and the effect
of the solar zenith angle have also been investigated.

We have found that for all conditions the SDR trend is stronger in
winter (\(0.7\), \(0.4\) and \(0.76\,\%/\text{year}\), for all- clear-
and cloudy skies respectively). For all- and cloudy-skis, in autumn
there are the next larger trends (\(0.42\) and \(0.19\,\%/\text{year}\),
respectively), the rest cases bare a negligable seasonal trend. The SDR
trend is increasing with increasing solar zenith angle (SZA) for
all-skies, from about \(0.2\) to \(0.6\,\%/\text{year}\). For clear- and
cloudy-skies remains low, about \(0.1\) --
\(0.2\,\%/\text{year}\,\%/\text{year}\) for all SZAs below
\(\sim 75^\circ\), and rizes rapidly for the larger SZAs.

Finally, we are discussing shorter-term variations in SDR anomalies by
examining the patterns of the cumulative sums of monthly anomalies from
the climatological mean, both before and after removing the long-term
trend.}

@@

The shortwave downward solar irradiance (SDR) at Earth's surface plays a
significant role, on its climate. Changes of the SDR can be related to
changes on Earth's energy budget, the mechanisms of climate change, and
water and carbon cycles \citep{Wild2009}. It can also affect solar and
agricultural production, and all living organisms. Studies of SDR
variability, have identified some distinct SDR trends on different
regions of the world on different time periods. The term `brightening'
is generally used to describe periods of positive SDR trend, and
`dimming' for negative trend \citep{Wild2009}. There are many cases in
the long term records of irradiance, showing a systematic change in the
magnitude of the trend, occurring roughly in the last decades of the
20th century. On multiple stations in China, a dimming period was
reported until about 2000, followed by a brightening period
\citep{Yang2021}. A similar pattern was identified, with the breaking
point around 1980, for stations in Central Europe \citep{Wild2021} and
Brazil \citep{Yamasoe2021}. On global scale, artificial Intelligence
aided spatial analysis on continental level with data from multiple
stations reach similar conclusions for these regions and for the global
trend \citep{Yuan2021}.

@@

Due to the significant spatial and temporal variability of the trends,
and the contributing factors, there is a constant need to monitor and
investigate SDR in different sites in order to estimate the degree of
variability, and its relation to the local conditions. In this study, we
examine the trends of SDR, using ground-based measurements at
Thessaloniki, Greece, for the period 1993 to 2023, as derived from a
CM-21 pyranometer. We re-evaluated and extended the dataset used by
\citet{Bais2013}, we applied a different algorithm for the
identification of clear-/cloud-sky instances
\citep{Reno2016, Reno2012a}, and we derived the SDR trends for the
period of study, under different sky conditions (all-sky, clear-sky and
cloud-sky). Finally, we investigated the dependence of the trends on
solar zenith angle and season.

@@

The SDR data were measured with a Kipp \& Zonen CM-21 pyranometer
operating continuously at the Laboratory of Atmospheric Physics of the
Aristotle University of Thessaloniki (\(40^\circ\,38'\,\)N,
\(22^\circ\,57'\,\)E, \(80\,\)m~a.s.l.) since 1993. Here we use data for
the period from 1993-04-13 to 2023-04-13. The monitoring site is located
near the city centre thus we expect that measurements are affected by
the urban environment. During the study period, the pyranometer has been
independently calibrated three times at the Meteorologisches
Observatorium Lindenberg, DWD, verifying that stability of the
instrument's sensitivity is better than \(0.7\%\) relative to the
initial calibration by the manufacturer. Along with SDR, the direct beam
radiation (DNI) is also measured with a collocated Kipp \& Zonen CHP-1
pyrheliometer since 2016-04-01. The DNI data were used as auxiliary data
to support the selection of appropriate thresholds in the clear sky
identification algorithm (CSid), which is discussed later. It is noted
that the limited dataset of DNI was not used for the identification of
clear sky cases in the entire SDR series to avoid any selection bias due
to unequal length of the two datasets. There are four distinct steps in
the creation of the dataset analysed here: a)~the acquisition of
radiation measurements from the sensors, b)~the data quality check,
c)~the identification of ``clear sky'' conditions from the radiometric
data, and d)~the aggregation of data and trend analysis.

@@

In order to estimate the effect of the sky conditions on the long term
variability of SDR, we created three datasets by characterizing each
one-minute measurement with a corresponding sky-condition flag (i.e.,
all-sky, clear-sky and cloudy-sky). To identify the clear-cases we used
the method proposed by \citet{Reno2016}, which requires the definition
of some site specific parameters. These parameters, were produced by an
iterative process, as the original authors proposed, and are discussed
in the next section.

We note that all methods have some subjectivity in the definition of
clear or cloudy sky cases. As a result, the details of the definition
are site specific, and they rely on a combination of thresholds and
comparisons with ideal radiation models and statistical analysis of
different signal metrics. The CSid algorithm was calibrated with the
main focus, to identify the presence of clouds. Despite the fine-tuning
of the procedure, in a few marginal cases false positive or false
negative results were identified by manual inspection. However, due to
their small number, they cannot affect the final results of the study.
For completeness, we provide below a brief overview of the clear-sky
identification algorithm (CSid), along with the site specific
thresholds.

@@

The linear trends were calculated using a first order autoregressive
model with lag of 1 day with fitting method of the ``maximum
likelihood'' \citep{Gardner1980, Jones1980}, by implementing the
function ``arima'' from the library ``stats'' of the R programming
language \citep{RCT2023}. The trends were reported together with the
\(2\sigma\) errors (Table~\ref{tab:trendtable}).

@@

We calculated the linear trends of SDR, from the departures of the mean
daily values from the daily climatology and for the three sky conditions
(Table~\ref{tab:trendtable}). In Figure~\ref{fig:trendALL} we present
only the time series under all-sky conditions; the plots for clear-sky
and cloud-sky conditions, are shown in the Appendix
(Figures~\ref{fig:trendCLEAR} and~ \ref{fig:trendCLOUD}). In the studied
period, there is no significant break or change in the variability
pattern of the time series. The linear trends in all three datasets are
positive and around \(0.4\,\%/y\) for all-sky and cloudy-sky conditions,
while for clear-skies the trend is much smaller
(\textasciitilde{}\(0.1\,\%/y\)). The linear trends were calculated
taking into account the autocorrelation of the time series and all three
are statistically significant at least at the \(95\,\%\) confidence
level, as they are larger than the corresponding \(2\sigma\) errors. The
clear-sky trend is very small suggesting a small effect from aerosols
and water vapour which are the dominant factors of the SDR variability.
In contrast the large positive trend of SDR under cloudy skies can be
attributed to reduction in cloud cover and/or cloud optical depth. Lack
of continuous cloud observations that could support these findings does
not allow drawing firm conclusions. However, there are indications that
the total cloud-cover as inferred from the ERA5 analysis for the grid
point of Thessaloniki is decreasing over the period of study.

@@

Although the year-to-year variability of the anomalies (Figure
\ref{fig:trendALL} and Figures~\ref{fig:trendCLEAR},
\ref{fig:trendCLOUD} in Appendix), shows a rather homogeneous behaviour,
plots of the cumulative sums (CUSUM) \citep{Regier2019} of the anomalies
can reveal different structures in the records of all three sky
conditions. For time series with a uniform trend, we would expect the
CUSUMs of the anomalies to have a symmetric `V' shape centred around the
middle of the series. This would indicate that the anomalies are evenly
distributed around the climatological mean, and for a positive uniform
trend, the first half is below and the second half above the
climatological mean. In our case, there is a more complex evolution of
the anomalies. For all-skies (Figure~\ref{fig:cusummonth-1}), we observe
three rather distinct periods: a) a downward part between the start of
the datasets and about 2000, denoting that all anomalies are negative
thus below the climatology; b) a relatively steady part lasting for
almost 20 years suggesting little variability in SDR anomalies; and c) a
steep upward part to present indicating anomalies above the climatology.
The CUSUMs for cloudy-skies (Figure~\ref{fig:cusummonth-3}), show a
similar behaviour with some short-term differences that do not change
the overall pattern. For clear skies (Figure~\ref{fig:cusummonth-2}), a
monotonic downward tendency is evident until 2004, suggesting that the
anomalies are all negative. After 2004 the anomalies turn to positive at
a fast rate for about five years and at a slower rate thereafter.

@@

The trends under clear- and cloudy-sky conditions are of the same
direction, and it would be expected that their sum is similar to the
all-sky trend. This does not happen, especially for winter, likely due
to the way the monthly means for clear and cloudy skies were calculated.
Daily means were calculated only when at least \(60\,\%\) of the clear-
or cloudy-sky data were available (see sect.
\ref{aggregationstatistical}).

@@

We have analysed a 30-year dataset of SRD measurements in Thessaloniki
Greece (1993 -- 2023) aiming to identify the long term variability of
solar irradiance under different sky conditions. Under all-sky
conditions there is a positive trend in SDR of \(0.38\,\%/y\)
(brightening). A previous study \citep{Bais2013} for the period 1993 --
2011 reported also a positive trend, of \(0.33\,\%/y\). The slight
increase of this trend indicates that the brightening of SDR continues
and is probably caused by continuing decreases in aerosol optical depth
and the optical thickness of clouds over the area. A smaller trend has
been found under clear-sky conditions (\(0.097\,\%/y\)) which supports
the notion that part of the brightening is caused by decreasing
aerosols. \citet{Siomos2020} have shown that aerosol optical depth over
Thessaloniki is decreasing constantly at least up to 2018.
Unfortunately, for this study aerosol data for the entire period were
not available in order to quantify their effect on SDR. The attenuation
of SDR by aerosols over Europe have been proposed as major factor by
\citet{Wild2021}. The brightening effect on SDR under cloudy-sky
conditions (\(0.41\,\%/y\)), suggests that cloud optical thickness is
also decreasing during this period. As long term data of cloud optical
thickness are also not available for the region, we cannot draw
quantitative conclusions. \citet{Lozano2023} reported a Cloud Radiative
Forcing (CRF) SDR trend of \(1.22\,W/m^2/y\) for the city of Granada
Spain (\(37^\circ\,10'\,\)N, \(3^\circ\,37'\,\)E, \(680\,\)m~a.s.l.). We
found a comparable result, for the same parameter, using the long term
trend of all- and clear-skies, as follows
\(\text{CRF} = 1.46 - 0.501 = 0.959\,W/m^2/y\).

### 2023-12-20 appendix.tex
Commit 2023-12-19 17:11
Commit 2023-12-19 18:21
Commit 2023-12-20 11:14

\begin{figure}[h!]
    {\centering
        \includegraphics[width=0.7\linewidth]{./images/LongtermTrends-7}

    }
    \caption{Anomalies (\%) of the daily clear-sky SDR, relative to climatological values for1993 - 2023. The black line shows the long term linear trend for clear-sky conditions.}\label{fig:trendCLEAR}
\end{figure}

\begin{figure}[h!]
    {\centering
        \includegraphics[width=0.7\linewidth]{./images/LongtermTrends-10}

    }
    \caption{Anomalies (\%) of the daily cloud-sky SDR, relative to climatological values for 1993 - 2023. The black line shows the long term linear trend for cloud-sky conditions.}\label{fig:trendCLOUD}
\end{figure}

>>>>

\begin{figure}[h!]
    {\centering
        \includegraphics[width=0.75\linewidth]{./images/LongtermTrends-2}

    }
    \caption{Anomalies (\%) of the daily clear-sky SDR, relative to climatological values for 1993 -- 2023. The black line shows the long term linear trend for clear-sky conditions.}\label{fig:trendCLEAR}
\end{figure}

\begin{figure}[h!]
    {\centering
        \includegraphics[width=0.75\linewidth]{./images/LongtermTrends-3}

    }
    \caption{Anomalies (\%) of the daily cloud-sky SDR, relative to climatological values for 1993 -- 2023. The black line shows the long term linear trend for cloud-sky conditions.}\label{fig:trendCLOUD}
\end{figure}
